---
title: "Agreement coefficients, correlations, inference, and machine learning"
subtitle: "Advances in motometry"
institute: "Department of Data Science and Analytics, BI Norwegian Business School"
author: Jonas Moss
date: 9-20-2024
execute:
  echo: true
format:
  revealjs: 
    theme: [default]
    width: 1600
    height: 900
    embed-resources: True
    mainfont: "Verdana"
    min-scale: 0.2
    max-scale: 2
    echo: true
reference-location: margin
citation-location: margin
fig-cap-location: top
lightbox: true
---

```{r}
#| echo: False
library("ISLR")
Auto = Auto[, 1:8]
```

## Motometry

```{r}
head(Auto)
```

. . .

Let's predict `mpg` (miles per gallon).

```{r}
mod1 = lm("mpg ~ weight * cylinders + year", data = Auto[1:300,])
mod2 = lm("mpg ~ weight * cylinders * year", data = Auto[1:300,])
```

. . .

The adjusted $R^2$ of both models are similar.

```{r}
summary(mod1)$adj.r.squared
summary(mod2)$adj.r.squared
```

## How are the models different?

```{r}
#| echo: False
#| fig-align: center
x = predict(mod1, Auto[301:392,])
y = predict(mod2, Auto[301:392,])
plot(predict(mod1, Auto[301:392,]), predict(mod2, Auto[301:392,]), xlab = "Model 1 prediction", ylab = "Model 2 predictions")
abline(a = 0, b=1)
abline(lm(y ~ x), col = "blue", lwd=2)
```

. . .

Seems like `Model 2` predicts smaller `mpg` when `Model 1` predicts small `mpg`, and higher when it predicts high `mpg`s.

**How much do the models agree?** 

## A clever trick?

* Can't use correlations - we want to check deviations from the line $y=x$, not $y = a+bx$!
* Two models can have correlation $1$ disagree completely (think $b = -1$.)

. . .


#### Possible solution

:::: {.columns}

:::{.column}

Plotting `c(x,y)` vs `c(y, x)` when `x`, `y` are the predictions of both models.

```{r}
#| echo: False
#| fig-align: "center"
x = predict(mod1, Auto[301:392,])
y = predict(mod2, Auto[301:392,])
plot(x, y, xlab = "x", ylab = "y")
points(y, x, col = "red")
abline(a = 0, b=1)
abline(lm(I(c(x,y)) ~ I(c(y,x))), col = "blue", lwd=2)
legend("topleft", legend = c("x: Model 1 predictions, y: Model 2 predictions", "x: Model 2 predictions, y: Model 1 predictions"), bty = "n", col = c("black", "red"),, pch= c(1,1))
```

:::

:::{.column}
:::{.fragment}

Calculating the correlation.
```{r}
x = predict(mod1, Auto[301:392,])
y = predict(mod2, Auto[301:392,])
cor(c(x,y), c(y, x))
```

Looks *ad hoc* and stupid? Sure!

**But it works!**

:::
:::

:::: 

## The approach is sound

Define the *quadratic kappa* (Fleiss, 1973) as $$\kappa_F(X,Y) = 1- \frac{E[(X-Y)^2]}{E[(X'-Y')^2]},$$ where $X',Y'$ are independent samples from their marginal distribution.

. . .

Then
$$\textrm{Cor}(X\frown Y, Y\frown X) = \kappa_F(X,Y),$$
where $X\frown Y$ and $Y\frown X$ are defined so that its sample estimator is `cor(c(x,y), c(y,x))`.

**You can estimate agreement using this bonkers method!**

## Inference
**Naïve approach.** Use standard OLS theory. This is an OLS, rigth?

. . . 

**[NO!]{style="color:red;"}** The observations are dependent in an insane way. OLS theory won't help us.

. . .

* Instead, we do it the "SEM way", using the $\Gamma$ matrix and so on.

* Exactly the same results can be obtained using the theory of $U$-statistics.

. . .

***But first, let's look at the case of multiple raters (models)!***



# Multiple raters (or models)
## Agreement for multiple raters

Say we have $n$ models / raters. How do we measure agreement at **one** prediction? 

. . .

### Sample variance

::: {.incremental}
* $\textrm{var}(x_1, x_2, x_3, \ldots, x_n) = \frac{\overline{(x_i-\overline{x})^2}}{n}.$
* Measures the mean squared deviation from the "center".
* Can be generalized to any loss function, with inferential accompanying theory. (Moss, 2024)
:::

::: {.fragment}

### Second approach

* Measure $(x_i - x_j)^2$ for all pairs $(i,j)$.
* These approaches are *equivalent*, but only for the square loss.

:::

## Chance-correct measures of agreement

$$\kappa_F = 1 - \frac{D}{D_C},$$

* $D$ is expected disagreement under the true joint distribution of the ratings
* $D_C$ is the expected disagreement when all ratings are independently distributed from their marginal distributions.

. . .

:::: {.columns}

::: {.column}
**Variance form.**

* $D=E_P[\textrm{var}(X_1,X_2,\ldots, X_n),$
* $D_C = E_{\overline{P}}[\textrm{var}(X_1,X_2,\ldots, X_n),$ where $\overline{P}$ is the product measure of the marginals.
:::

::: {.column}
**Sum form.**

* $D = \sum_{i>j} E_P((X_i - X_j)^2)$
* $D = \sum_{i>j} E_P((X_i - X_j')^2),$ where $X_j$ is independently distributed of $X_i$ according to its marginal distribution.
:::

::::

## Example

Let's fit four models.
```{r}
mod1 = lm("mpg ~ weight * cylinders + year", data = Auto[1:300,])
mod2 = lm("mpg ~ weight * cylinders * year", data = Auto[1:300,])
mod3 = lm("mpg ~ weight * cylinders ", data = Auto[1:300,])
mod4 = lm("mpg ~ weight + year", data = Auto[1:300,])
```

. . . 
```{r}
#| echo: False
#| fig-align: "center"
x = predict(mod1, Auto[301:392,])
y = predict(mod2, Auto[301:392,])
z = predict(mod3, Auto[301:392,])
w = predict(mod4, Auto[301:392,])
plot(x, y, xlab = "Model 1 predictions", ylab = "Model x predictions")
points(x, z, col = "red")
points(x, w, col = "blue")
abline(a = 0, b=1)
legend("topleft", legend = c("Model 2", "Model 3", "Model 4"), bty = "n", col = c("black", "red", "blue"),, pch= c(1,1))
```

Pretty poor agreement between, e.g., model 1 and model 3 here!

## Population definition
Let $X$ be a random vector with mean $\mu$ and covariance $\Sigma$. The quadratic Fleiss' kappa is then

$$\kappa_F = \frac{1^{T}\Sigma1-\textrm{tr}\Sigma-R\left(\overline{\mu^{2}}-\overline{\mu}^{2}\right)}{(R-1)\textrm{tr}\Sigma+(R-1)R\left(\overline{\mu^{2}}-\overline{\mu}^{2}\right)}.$$

::: {.incremental}
* The standard estimator replace $\mu$ and $\Sigma$ with their (biased) sample counter-parts.
* **Inference?** A clear-cut example of standard asymptotic theory involving the fourth-order matrix $\Gamma$ in addition to $\mu$ and $\Sigma$.
:::

## The correlation definition still holds!

Let's construct some predictions for each model.

```{r}
x = predict(mod1, Auto[301:392,])
y = predict(mod2, Auto[301:392,])
z = predict(mod3, Auto[301:392,])
w = predict(mod4, Auto[301:392,])
```

. . .

I have dutifully implemented $\kappa_F$ in an `R` package.
```{r}
quadagree::fleiss(cbind(x,y,z,w))
```

. . .

But, behold!

```{r}
x1 = c(x,x,x,y,y,y,z,z,z,w,w,w)
x2 = c(y,z,w,x,z,w,x,y,w,x,y,z)
cor(x1, x2)
```

. . .

* A precise formulation of the relationship is tedious.
* We just need to make sure every pair `x`, `y` appears twice when calculating the correlation.

## Missing values

Assume ratings are missing at random. Let $S$ and $\hat{\mu}$ be calculated using pairwise available data.

. . . 

From van Praag et al. (1985) we know that
$$\sqrt{n}(\textrm{vec}S - \textrm{vec}\Sigma) \stackrel{d}{\to} N(0, \Gamma \odot \Pi),$$ 

. . .

Moreover, we have
$$\sqrt{n}(\hat{\mu} - \mu) \stackrel{d}{\to}  N(0, \Sigma + (P-I) \textrm{diag}(\Sigma)).$$
Here $P$ and $\Pi$ encodes the probability of missingness.

Everything else is the delta method. (Tedious!)


## Four models with missing values

We are not always able to obtaining all ratings, just a random subset.
```{r}
#| echo: False
noise <- \() sample(c(0,NA), size=92, replace = TRUE, prob = rep(0.5, 2))
x_ = x + noise()
y_ = y + noise()
z_ = z + noise()
w_ = w + noise()
```

```{r}
#| eval: False
x, y, z, w # original predictions from 4 models
x_, y_, z_, w_ # predictions with missing data 
```

. . .

```{r}
quadagree::fleissci(cbind(x_,y_,z_,w_))
```

. . . 
 
```{r}
quadagree::fleissci(cbind(x,y,z,w))
```

## Existing methods are incorrect

* Existing methods use ad-hoc weighting methods introduced by Gwet (2020). 

* These weighting methods are inconsistent, with asymptotically incorrect level, and do not use all available data.

```{r}
data <- cbind(x_,y_,z_,w_)
data <- data[rowSums(is.na(data)) <= 1, ]
#data <- round(data)
irrCAC::fleiss.kappa.raw(data, weights = "quadratic")$est
```
* Missing data are everywhere when measuring agreement.
* Consistency shoul be the *bare minimum* requirement for an estimator.

## Summary

1. Quadratic kappa is a correlation coefficient.
2. We can do inference with quadratic kappa with missing values.
3. Quadratic kappa with multiple raters can be interpreted in several ways.
4. `quadagree` implements the estimation and inference methods.
5. Current methods for handling missing data are **incorrect**. 

. . .

Moreover,

1. These methods show great potential for motometry and other machine learning applications.
2. Preprint out in 1-3 months I think. 
3. Will be submitted to Motometrika.
